---
title: "GGplot"
author: "Anna Lorenc"
date: "10/16/2017"
output: pdf_document
---





First, let's prepare a play dataset. The package `ggplot` comes with in-buit dataset diamonds. As it is very big, we'll prepare a subset `diamonds_small` by randomly choosing 4000 entries.


```{r ,  eval=FALSE}
diamonds
diamonds <- data.frame(diamonds)
dim(diamonds)

diamonds_small <-diamonds[sample(1:nrow(diamonds), 4000),]

```

#General syntax:

  `ggplot(data=<DATA>) + <GEOM_FUNCTION>(mapping=aes(<MAPPINGS>))`


```{r ,  eval=FALSE}
ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price))

ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price, colour=cut)) # notice the legend

ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price, colour=color)) 

ggplot(data = diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price, colour=color, shape=cut))


```

  
Available aesthetics depend on the geom. Most used:

* x, y,
* col, fill,  alpha,
* linetype, linewidth,
* shape, size.


Instead of specyfying aesthetics by mapping, we can use a direct assignement:

```{r,  eval=FALSE}
ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price), colour="red")

ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price), shape=3 )

ggplot(data=diamonds_small ) +
  geom_point(mapping=aes(x=carat, y=price), shape=3, size=3 )

```

Plot might be directly printed or stored in an object. This object might be printed on screen or into a file. It might be also modified and combined with other plots.

```{r,  eval=FALSE}
g_diamonds_col <- ggplot(data=diamonds_small ) + 
  geom_point(mapping=aes(x=carat, y=price, colour=color))

g_diamonds_col 

ggsave(g_diamonds_col, file="g_diamonds_col.pdf")

ggsave(g_diamonds_col, file="g_diamonds_col.pdf", width=22, height = 20, units = "cm")

```
 
 
### Some other geoms

##### Boxplot 
```{r,  eval=FALSE}
ggplot(data=diamonds) + 
  geom_boxplot(mapping=aes(x=cut, y=price))

ggplot(data=diamonds) + 
  geom_boxplot(mapping=aes(x=cut, y=price, colour=cut))

ggplot(data=diamonds) + 
  geom_boxplot(mapping=aes(x=cut, y=price, fill=cut))

ggplot(data=diamonds) + 
  geom_boxplot(mapping=aes(x=cut, y=price, fill=color))

ggplot(data=diamonds) + 
  geom_boxplot(mapping=aes(x=cut, y=price, fill=color), notch=TRUE)
```

##### Violin plot 
```{r, eval=FALSE}
ggplot(data=diamonds) + 
  geom_violin(mapping=aes(x=cut, y=price))

```

##### Histogram
```{r, eval=FALSE}
ggplot(data=diamonds) + 
  geom_histogram(mapping=aes(x=price, fill=cut), col="black")

```


##### Fitted line 
By default a loess fit with confidence intervals, might be changed.
```{r, eval=FALSE}
ggplot(data=diamonds) + 
  geom_smooth(mapping=aes(y=price, x=carat))

ggplot(data=diamonds) + 
  geom_smooth(mapping=aes(y=price, x=carat, col=cut))
```

And ~30 other geoms.

### Combining layers
A geom defines one layer of a plot, we can combine layers:

```{r,  eval=FALSE}
ggplot(data=diamonds_small) +
  geom_point(mapping=aes(x=carat, y=price, col=cut) ) +
  geom_smooth(mapping=aes(y=price, x=carat, col=cut)) 
```


### Plot fine-tuning:


* setting plot limits
* changing axes
* log-scale
* adding title and axis names
* changing tick names
* removing plot elements(ticks, axes, legend)
* manually defining colours, shapes etc.

```{r, echo=FALSE, eval=FALSE} 
 ggplot(data=diamonds_small) +
  geom_point(mapping=aes(x=carat, y=price, colour=color, shape=cut), size=4) +
  xlim(1.5, 3) +
  ylim(10000, 20000)

ggplot(data=diamonds_small) +
  geom_point(mapping=aes(x=carat, y=price, colour=color, shape=cut), size=4) +
  xlim(1.5, 3) +
  ylim(10000, 20000)+
  coord_flip()

```


### Facetting 

Splitting data into facets by categorical variables (one or more)

```{r,  eval=FALSE}
ggplot(data=diamonds_small) +
  geom_point(mapping=aes(x=carat, y=price, col=cut) )+
  facet_wrap(~clarity)
```


Package `cowplot` for arranging several plots into figures (think: histogram on top of a scatterplot). 
