---
title: "Introduction to R and RStudio part2 - challenges"
author: "Anna Lorenc anna.lorenc@kcl.ac.uk"
source: Rmd
output: pdf_document
---

##### Challenge 1
What do you think the following will produce?

```{r}
quiz_vector <- c(2,6,'3')
```

##### Challenge 2
 Create a list of length two containing a character vector for each of the sections in this part of the workshop:

 - Data types
 - Data structures

 Populate each character vector with the names of the data types and data
 structures we've seen so far.

###Sunshine data
Data is in the location:

`https://raw.githubusercontent.com/Ania-varia/R_for_immunologists/master/blob/data/`

File of interest:
`sunshine.csv`


##### Challenge 3

Given the following code:
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
```
Write a subsetting command to return the values in x that are greater than 4 and less than 7.

###GapMinder data
Data is in the location:

`https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/`

File of interest:
`gapminder-FiveYearData.csv`

##### Challenge 4

Extract all rows with information about Montenegro  from the Gapminder data.

##### Challenge 5

Selecting elements of a vector that match any of a list of components
is a very common data analysis task. For example, the gapminder data set
contains `country` and `continent` variables, but no information between
these two scales. Suppose we want to pull out information from southeast
Asia: how do we set up an operation to produce a logical vector that
is `TRUE` for all of the countries in southeast Asia and `FALSE` otherwise?

Define SE Asia as:
```{r, eval=FALSE}
seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
```

There's a wrong way (using only `==`), which will give you a warning;
a clunky way (using the logical operators `==` and `|`); and
an elegant way (using `%in%`). See whether you can come up with all three
and explain how they (don't) work.

##### Challenge 6

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

```{r, eval=FALSE}
gapminder[gapminder$year = 1957,]
```

2. Extract all columns except 1 through to 4

```{r, eval=FALSE}
gapminder[,-1:4]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r, eval=FALSE}
gapminder[gapminder$lifeExp > 80]
```

4. Extract the first row, and the fourth and fifth columns
  (`lifeExp` and `gdpPercap`).

```{r, eval=FALSE}
gapminder[1, 4, 5]
```

5. Advanced: extract rows that contain information for the years 2002
and 2007

```{r, eval=FALSE}
gapminder[gapminder$year == 2002 | 2007,]
```


##### Challenge 7
Create a new `data.frame` called `gapminder_small` that only contains rows 1 through 9 and 19 through 23. You can do this in one or two steps.

##### Challenge 8
Plot the population of Senegal in time.

##### Challenge 9
Calculate the average income for countries on the continent of Asia.


  
## Common R functions
### Creating and saving data

*  _c(),vector()_ create a vector
*  _matrix()_ create a matrix
*  _data.frame()_ create a data frame
*  _list()_ create a list
*  _cbind(), rbind()_ add columns/rows to matrix/data frame


### Text and data processing

*  _paste()_ paste character strings
*  _sub()_ substitute in character strings
*  _sort()_ sort elements
*  _order()_ give order of elements
*  _t()_ transpose a matrix
*  _apply()_ perform a function column or row-wise


### Accessing object elements

*  _vector[index]_ index-th element of a vector
*  _vector[elementName]_ element of a vector named elementName
*  _object[rows, columns]_ elements in rows and columns of a matrix/dataframe
*  _object[rownames, columnnames]_ elements in rows and columns of a matrix/dataframe
*  _list[index]_ - output is a list
*  _list[[index]]_ - output is a list element
*  _list[[name]]_ or list$name - output is a list element
*  _names(object)_ names of an object
*  _colnames(object),rownames(object)_ column or row names of an object


### Inspecting an object

*  _length()_ length of an object
*  _dim()_ dimensions  of an object
*  _ncol(), nrow()_ number of columns or rowsof an object
*  _head(), tail()_ show beginning or end of an object
*  _str()_ structure of an object
*  _class()_ class of an object
*  _typeof() type of data
*  _is.character(),is.numeric()_ check if object is of type character or numeric
*  _as.character(), as.numeric(), as.data.frame(), as.factor()_ force an object into type character or numeric or factor, into class data.frame

### Statistics

*  _sum()_ arithmetic sum
*  _mean()_ mean
*  _sd()_ standard deviation
*  _min(), max(), range(), summary()_ minimum, maximum or both
*  _which.min(), which.max()_ which element of an object is min or max
*  _rnorm()_ draw from a normal distribution
*  _rowSums(), colSum()s, rowMeans(), colMeans()_ perform summing or average by rows or columns
*  _t.test()_ perform t-test (see also prop.test(), chisq.test(), wilcox.test())
*  _cor.test()_ correlation
*  _rnorm()_ -draw from a nomal distribution

### Plots

*  _plot()_ plot, type of plot depends on data
*  _barlot()_ 
*  _boxplot()_
*  _hist()_ histogram
*  _abline()_ add a line to an existing plot
*  _text()_ add text labels to an existing plot
*  _legend()_ add legend to an existing plot

### Workspace and environment

*  _ls()_ show objects in the workspace
*  _rm()_ remove objects
*  _read.table()_ read data from a file into a data frame
*  _write.table()_ output a data frame into a file
*  _save()_ save objects into .Rdata file
*  _load()_ load objects from a .Rdata file
*  _getwd()_ print working directory
*  _setwd()_ set working directory
*  _list.files()_ list files from a directory
*  _install.package()_ install a package
*  _library(), require()_  load a package
