---
title: "Introduction to R and RStudio part 2"
author: "Anna Lorenc anna.lorenc@kcl.ac.uk"
source: Rmd
output: pdf_document
---

## Data Structures

One of R's most powerful features is its ability to deal with tabular data -
such as you may already have in a spreadsheet or a CSV file. Let's start by
making a toy dataset in your `data/` directory, called `feline-data.csv`:

```{r, eval=FALSE , results='hide'}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

> ####Tip: Editing Text files in R
>
> Alternatively, you can create `data/feline-data.csv` using a text editor,
> or within RStudio with the **File -> New File -> Text File** menu item.


```{r, echo = FALSE, results='hide' , results='hide'}
cats.df <- data.frame(coat = c("calico", "black", "tabby"),
                      weight = c(2.1, 5.0, 3.2),
                      likes_string = c(1, 0, 1))

write.csv(cats.df, "feline-data.csv", row.names = FALSE)
```

We can load this into R via the following:

```{r  , results='hide'}
cats <- read.csv(file = "feline-data.csv")
cats
```


We can begin exploring our dataset right away, pulling out columns by specifying
them using the `$` operator:

```{r , results='hide'}
cats$weight
cats$coat
```

We can do other operations on the columns:

```{r , results='hide'}
## Say we discovered that the scale weighs two Kg light:
cats$weight + 2
paste("My cat is", cats$coat)
```

But what about:

```{r, eval=FALSE , results='hide'}
cats$weight + cats$coat
```

Understanding what happened here is key to successfully analyzing data in R.

## Data Types

If you guessed that the last command will return an error because `2.1` plus
`"black"` is nonsense, you're right - and you already have some intuition for an
important concept in programming called *data types*. We can ask what type of
data something is:

```{r , results='hide'}
typeof(cats$weight)
```

There are 5 main types: `double`, `integer`, `complex`, `logical` and `character`.

```{r , results='hide'}
typeof(3.14)
typeof(1L) # The L suffix forces the number to be an integer, since by default R uses float numbers
typeof(1+1i)
typeof(TRUE)
typeof('banana')
```

No matter how complicated our analyses become, all data in R is interpreted as one of these
basic data types. This strictness has some really important consequences.

A user has added details of another cat. This information is in the data frame cats_bad:


```{r , results='hide'}
cats_bad <- data.frame(coat = factor(c("calico", "black", "tabby", "tabby")),
                   weight = factor(c(2.1, 5.0, 3.2, '2.3 or 2.4')),
                   likes_string = c(1, 0, 1, 1))
```

Check what type of data we find in the
`weight` column:

```{r , results='hide'}
cats_bad
typeof(cats_bad$weight)
```

Oh no, our weights aren't the double type anymore! If we try to do the same math
we did on them before, we run into trouble:

```{r , results='hide'}
cats$weight + 2
```

What happened? Within a data.frame each column may be of different type, but within a column type has to be the same.
R insists that everything in a column be the same basic type; if it can't understand
*everything* in the column as a double, then *nobody* in the column gets to be a
double. 


> 
>
> Why is R so opinionated about what we put in our columns of data?
> How does this help us?

> >
> > By keeping everything in a column the same, we allow ourselves to make simple
> > assumptions about our data; if you can interpret one entry in the column as a
> > number, then you can interpret *all* of them as numbers, so we don't have to
> > check every time. This consistency is what people mean when they talk about 
> > *clean data*; in the long run, strict consistency goes a long way to making 
> > our lives easier in R.


```{r , results='hide'}
str(cats$weight)
```

we see that `cats$weight` is a vector, too - *the columns of data in
data.frames are all vectors*, and that's the root of why R forces everything in
a column to be the same basic data type.


##### Challenge 1
What do you think the following will produce?

```{r , results='hide'}
quiz_vector <- c(2,6,'3')
```

This is something called *type coercion*, and it is the source of many surprises
and the reason why we need to be aware of the basic data types and how R will
interpret them. When R encounters a mix of types (here numeric and character) to
be combined into a single vector, it will force them all to be the same
type. Consider:

```{r , results='hide'}
coercion_vector <- c('a', TRUE)
coercion_vector
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```

The coercion rules go: `logical` -> `integer` -> `numeric` -> `complex` ->
`character`, where -> can be read as *are transformed into*. You can try to
force coercion against this flow using the `as.` functions:

```{r , results='hide'}
character_vector_example <- c('0','2','4')
character_vector_example
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

As you can see, some surprising things can happen when R forces one basic data
type into another! Nitty-gritty of type coercion aside, the point is: if your
data doesn't look like what you thought it was going to look like, type coercion
may well be to blame; make sure everything is the same type in your vectors and
your columns of data.frames, or you will get nasty surprises!

But coercion can also be very useful! For example, in our `cats` data
`likes_string` is numeric, but we know that the 1s and 0s actually represent
`TRUE` and `FALSE` (a common way of representing them). We should use the
`logical` datatype here, which has two states: `TRUE` or `FALSE`, which is
exactly what our data represents. We can 'coerce' this column to be `logical` by
using the `as.logical()` function:

```{r , results='hide'}
cats$likes_string
cats$likes_string <- as.logical(cats$likes_string)
cats$likes_string
```

We said that columns in data.frames were vectors:

```{r , results='hide'}
str(cats$weight)
str(cats$likes_string)
```

These make sense. But what about

```{r , results='hide'}
str(cats$coat)
```

## Factors

Another important data structure is called a *factor*. Factors usually look like
character data, but are typically used to represent categorical information. For
example, let's make a vector of strings labelling cat colorations for all the
cats in our study:

```{r , results='hide'}
coats <- c('tabby', 'tortoiseshell', 'tortoiseshell', 'black', 'tabby')
coats
str(coats)
```

We can turn a vector into a factor like so:

```{r , results='hide'}
CATegories <- factor(coats)
class(CATegories)
str(CATegories)
```

Now R has noticed that there are three possible categories in our data - but it
also did something surprising; instead of printing out the strings we gave it,
we got a bunch of numbers instead. R has replaced our human-readable categories
with numbered indices under the hood, this is necessary as many statistical
calculations utilise such numerical representations for categorical data:

```{r , results='hide'}
typeof(coats)
typeof(CATegories)
```


```{r , results='hide'}
class(cats)
```



Is there a factor in our `cats` data.frame? what is its name?
Try using `?read.csv` to figure out how to keep text columns as character
vectors instead of factors; then write a command or two to show that the factor
in `cats` is actually a character vector when loaded in this way.

In modelling functions, it's important to know what the baseline levels are. This
is assumed to be the first factor, but by default factors are labelled in
alphabetical order. You can change this by specifying the levels:

```{r , results='hide'}
mydata <- c("case", "control", "control", "case")
factor(mydata)
factor_ordering_example <- factor(mydata, levels = c("control", "case"))
str(factor_ordering_example)
```

In this case, we've explicitly told R that "control" should represented by 1, and
"case" by 2.

## Lists

Another data structure you'll want in your bag of tricks is the `list`. A list
is simpler in some ways than the other types, because you can put anything you
want in it:

```{r , results='hide'}
list_example <- list(1, "a", TRUE, 1+4i)
list_example
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE )
another_list
```

Slicing and accessing list elements. What are the outputs?
```{r  , results='hide'}
another_list[[1]]
another_list[1:2]
names(another_list) 
another_list[["title"]]
another_list[["myPatients"]] <- cats 
another_list

```


We can now understand something a bit surprising in our data.frame; what happens if we run:

```{r , results='hide'}
typeof(cats)
```

We see that data.frames look like lists 'under the hood' - this is because a
data.frame is really a list of vectors and factors, as they have to be - in
order to hold those columns that are a mix of vectors and factors, the
data.frame needs something a bit more flexible than a vector to put all the
columns together into a familiar table.  In other words, a `data.frame` is a
special list in which all the vectors must have the same length.

In our `cats` example, we have an integer, a double and a logical variable. As
we have seen already, each *column* of data.frame is a vector.

```{r , results='hide'}
cats$coat
cats[,1]
typeof(cats[,1])
str(cats[,1])
```

Each row is an *observation* of different variables, itself a data.frame, and
thus can be composed of elements of different types.

```{r , results='hide'}
cats[1,]
typeof(cats[1,])
str(cats[1,])
```




##### Challenge 2
 Create a list of length two containing a character vector for each of the sections in this part of the workshop:

 - Data types
 - Data structures

 Populate each character vector with the names of the data types and data
 structures we've seen so far.



## Solution to Challenge 2
```{r , results='hide'}
dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
answer <- list(dataTypes, dataStructures)
```


## Useful R functions

There are thousands of these - some in the basic R installation, some in additional packages. Let's try some of them.
First, download a file with data.
```{r  , results='hide'}
sunshine <- read.csv("https://raw.githubusercontent.com/Ania-varia/R_for_immunologists/master/blob/data/sunshine.csv")
str(sunshine)
sunshine
```

Let's compare something obvious with a statistical test already implemented in R. 

```{r, eval=FALSE , results='hide'}
# Here the function needs more than one argument (we compare two groups!)
t.test(x=London, y=Madrid, data =sunshine)
```

Result of a test might be assigned to an object. Let's do it and add yet another argument to the function

```{r, eval=FALSE , results='hide'}
exampleTtest <- t.test(x=sunshine$London, y=sunshine$Madrid, paired=TRUE)
exampleTtest
```

It is also possible to draw from distributions - here from normal, we need to specify mean and sd 
```{r , results='hide'}
# 
height <- rnorm(mean=164, sd=6, n=1000)
height
```


## Graphics 
R allows to make basic or very refined plots, especially with help of graphical packages like lattice/ggplot etc. 
Plots might be fine-tuned with additional parameters. Low-level graphical functions add elements like points, labels, axes, lines and legend.

```{r, eval=FALSE , results='hide'}


plot(sunshine$London,sunshine$Madrid)
plot(sunshine$London,sunshine$Madrid,xlim = c(20,370), ylim=c(20, 370))
plot(sunshine$London,sunshine$Madrid,xlim = c(20,370), ylim=c(20, 370),xlab = "London", ylab="Madrid")
plot(sunshine$London,sunshine$Madrid,xlim = c(20,370), ylim=c(20, 370),xlab = "London", ylab="Madrid", col="red", pch=19, main= "Monthly hours of sunshine")
abline(a = 0, b=1)
text(labels = month.name, sunshine$London,sunshine$Madrid,adj = c(1,1))

```



### Boxplot

```{r, eval=FALSE , results='hide'}
# boxplot - to compare sunny hours by location
boxplot(sunshine, col=2:3)
```

### Barplot

```{r ,eval=FALSE , results='hide'}
rownames(sunshine) <- month.name
sunshine
barplot(as.matrix(sunshine), beside=TRUE, col=1:12, main="Sunshine hours per month")
```

### Histogram 

```{r , eval=FALSE , results='hide'}
height <- rnorm(mean=164, sd=6, n=1000)
#Another very useful type of plot - histogram
hist(height)
# It is possible to fine-tune breaks by size or number
hist(height, breaks=100)
# Polishing the plot: proper title and axis names
hist(height, breaks=100, main= "Height of UK women", xlab= "cm") 
# adding elements: here a vertical line of the true mean
abline(v=164, col="red", lwd=2) #low level plotting function
# legend
legend(lwd=2, col="red", legend="actual mean", "topright")
```


### It is possible to plot straight to a file (.jpg, .pdf, .eps...)
```{r,  eval=FALSE  , results='hide'}


pdf("height_women_UK.pdf") # To start plotting to a file
  hist(height, breaks=100, main= "Height of UK women", xlab= "cm")
  abline(v=164, col="red", lwd=2)
  legend(lwd=2, col="red", legend="actual mean", "topright")
dev.off()                 #To close the file

```

## Working with data

Let's work with some more substantial dataset.
You can directly from R download a file from the internet:

```{r, eval=FALSE , results='hide'}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = "data/gapminder-FiveYearData.csv")
gapminder <- read.csv("data/gapminder-FiveYearData.csv")

```

You can read directly from excel spreadsheets without converting them to plain text first by using the [readxl](https://cran.r-project.org/web/packages/readxl/index.html) package.

The first thing we should always do is check out what the data looks like with `str`:

```{r, eval=FALSE , results='hide'}
str(gapminder)
```

We can also examine individual columns of the data frame with our `typeof` function:

```{r, eval=FALSE , results='hide'}
typeof(gapminder$year)
typeof(gapminder$country)
str(gapminder$country)
```

We can also interrogate the data frame for information about its dimensions;
remembering that `str(gapminder)` said there were 1704 observations of 6
variables in gapminder, what do you think the following will produce, and why?

```{r, eval=FALSE , results='hide'}
length(gapminder)
```


A fair guess would have been to say that the length of a data frame would be the
number of rows it has (1704), but this is not the case; remember, a data frame
is a *list of vectors and factors*

Now check for number of rows and columns and column names.


At this stage, it's important to ask ourselves if the structure R is reporting
matches our intuition or expectations; do the basic data types reported for each
column make sense? If not, we need to sort any problems out now before they turn
into bad surprises down the road, using what we've learned about how R
interprets data, and the importance of *strict consistency* in how we record our
data.

Once we're happy that the data types and structures seem reasonable, it's time
to start digging into our data proper. Check out the first few lines:


```{r, eval=FALSE , results='hide'}
head(gapminder)
```


Check how many different countries are in this data frame. Tip: unique() will list unique elements of a vector.

### Subsetting through logical operations

We can also use any logical vector to subset:

```{r , results='hide'}
x <-3:8
x
x[c(FALSE, FALSE, TRUE, FALSE, TRUE)]
```

Since comparison operators (e.g. `>`, `<`, `==`) evaluate to logical vectors, we can also
use them to succinctly subset vectors: the following statement gives
the same result as the previous one.

```{r , results='hide'}
x[x > 7]
```

Breaking it down, this statement first evaluates `x>7`, generating
a logical vector `c(FALSE, FALSE, TRUE, FALSE, TRUE)`, and then
selects the elements of `x` corresponding to the `TRUE` values.

We can use `==` to mimic the previous method of indexing by name
(remember you have to use `==` rather than `=` for comparisons):

```{r , results='hide'}
names(x)=letters[1:6]
x[names(x) == "d"]
```

####  Combining logical conditions

We often want to combine multiple logical
criteria. For example, we might want to find all the countries that are
located in Asia **or** Europe **and** have life expectancies within a certain
range. Several operations for combining logical vectors exist in R:

 * `&`, the "logical AND" operator: returns `TRUE` if both the left and right
   are `TRUE`.
 * `|`, the "logical OR" operator: returns `TRUE`, if either the left or right
   (or both) are `TRUE`.

You may sometimes see `&&` and `||` instead of `&` and `|`. These two-character operators
only look at the first element of each vector and ignore the
remaining elements. In general you should not use the two-character
operators in data analysis; save them
for programming, i.e. deciding whether to execute a statement.

 * `!`, the "logical NOT" operator: converts `TRUE` to `FALSE` and `FALSE` to
   `TRUE`. It can negate a single logical condition (eg `!TRUE` becomes
   `FALSE`), or a whole vector of conditions(eg `!c(TRUE, FALSE)` becomes
   `c(FALSE, TRUE)`).

Additionally, you can compare the elements within a single vector using the
`all` function (which returns `TRUE` if every element of the vector is `TRUE`)
and the `any` function (which returns `TRUE` if one or more elements of the
vector are `TRUE`).


#### Challenge 3

Given the following code:

```{r , results='hide'}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Write a subsetting command to return the values in x that are greater than 4 and less than 7.


#### Challenge 4

Extract all rows with information about Montenegro from the Gapminder data.

#### Challenge 5

Selecting elements of a vector that match any of a list of components
is a very common data analysis task. For example, the gapminder data set
contains `country` and `continent` variables, but no information between
these two scales. Suppose we want to pull out information from southeast
Asia: how do we set up an operation to produce a logical vector that
is `TRUE` for all of the countries in southeast Asia and `FALSE` otherwise?

Define SE Asia as:
```{r, eval=FALSE , results='hide'}
seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
```

There's a wrong way (using only `==`), which will give you a warning;
a clunky way (using the logical operators `==` and `|`); and
an elegant way (using `%in%`). See whether you can come up with all three
and explain how they (don't) work.



## Factor subsetting

Now that we've explored the different ways to subset vectors, how
do we subset the other data structures?

Factor subsetting works the same way as vector subsetting.

```{r , results='hide'}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
f[f %in% c("b", "c")]
f[1:3]
```

Skipping elements will not remove the level
even if no more of that category exists in the factor:

```{r , results='hide'}
f[-3]
```

#### Challenge 6

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

```{r, eval=FALSE , results='hide'}
gapminder[gapminder$year = 1957,]
```

2. Extract all columns except 1 through to 4

```{r, eval=FALSE , results='hide'}
gapminder[,-1:4]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r, eval=FALSE , results='hide'}
gapminder[gapminder$lifeExp > 80]
```

4. Extract the first row, and the fourth and fifth columns
  (`lifeExp` and `gdpPercap`).

```{r, eval=FALSE , results='hide'}
gapminder[1, 4, 5]
```

5. Advanced: extract rows that contain information for the years 2002
and 2007

```{r, eval=FALSE , results='hide'}
gapminder[gapminder$year == 2002 | 2007,]
```


#### Challenge 7
 Create a new `data.frame` called `gapminder_small` that only contains rows 1 through 9
 and 19 through 23. You can do this in one or two steps.

#### Challenge 8
Plot the population for Senegal in time 

#### Challenge 9
Calculate the average income for countries on the continent of Asia


  
## Common R functions

### Creating and saving data

*  _c(),vector()_ create a vector
*  _matrix()_ create a matrix
*  _data.frame()_ create a data frame
*  _list()_ create a list
*  _cbind(), rbind()_ add columns/rows to matrix/data frame


### Text and data processing

*  _paste()_ paste character strings
*  _sub()_ substitute in character strings
*  _sort()_ sort elements
*  _order()_ give order of elements
*  _t()_ transpose a matrix
*  _apply()_ perform a function column or row-wise


### Accessing object elements

*  _vector[index]_ index-th element of a vector
*  _vector[elementName]_ element of a vector named elementName
*  _object[rows, columns]_ elements in rows and columns of a matrix/dataframe
*  _object[rownames, columnnames]_ elements in rows and columns of a matrix/dataframe
*  _list[index]_ - output is a list
*  _list[[index]]_ - output is a list element
*  _list[[name]]_ or list$name - output is a list element
*  _names(object)_ names of an object
*  _colnames(object),rownames(object)_ column or row names of an object


### Inspecting an object

*  _length()_ length of an object
*  _dim()_ dimensions  of an object
*  _ncol(), nrow()_ number of columns or rowsof an object
*  _head(), tail()_ show beginning or end of an object
*  _str()_ structure of an object
*  _class()_ class of an object
*  _typeof()_ type of data
*  _is.character(),is.numeric()_ check if object is of type character or numeric
*  _as.character(), as.numeric(), as.data.frame(), as.factor()_ force an object into type character or numeric or factor, into class data.frame

### Statistics

*  _sum()_ arithmetic sum
*  _mean()_ mean
*  _sd()_ standard deviation
*  _min(), max(), range(), summary()_ minimum, maximum or both
*  _which.min(), which.max()_ which element of an object is min or max
*  _rnorm()_ draw from a normal distribution
*  _rowSums(), colSum()s, rowMeans(), colMeans()_ perform summing or average by rows or columns
*  _t.test()_ perform t-test (see also prop.test(), chisq.test(), wilcox.test())
*  _cor.test()_ correlation
*  _rnorm()_ -draw from a nomal distribution

### Plots

*  _plot()_ plot, type of plot depends on data
*  _barlot()_ 
*  _boxplot()_
*  _hist()_ histogram
*  _abline()_ add a line to an existing plot
*  _text()_ add text labels to an existing plot
*  _legend()_ add legend to an existing plot

### Workspace and environment

*  _ls()_ show objects in the workspace
*  _rm()_ remove objects
*  _read.table()_ read data from a file into a data frame
*  _write.table()_ output a data frame into a file
*  _save()_ save objects into .Rdata file
*  _load()_ load objects from a .Rdata file
*  _getwd()_ print working directory
*  _setwd()_ set working directory
*  _list.files()_ list files from a directory
*  _install.package()_ install a package
*  _library(), require()_  load a package
