---
title: "R for Immunologists:  working with FCS files"
author: "Anna Lorenc"
date: "02/11/2017"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, include=FALSE}
require(flowCore)
require(flowWorkspace)
require(gridExtra)
require(CytoML)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
dir_with_FCS_files <- "/Users/lorenca/Documents/teaching/RCourseImmunologist/immunology_specific/rawFCSfiles/"
#MY PATH TO WHERE THIS FILE IS ~/Documents/teaching/RCourseImmunologist/immunology_specific/FlowJo Workspace
```


###Reading in a .FCS file
Read in a single file into a `flowFrame`.
```{r, eval=FALSE}

require(flowCore)
dir_with_FCS_files <- PATH TO DIR WITH FCS FILE 
setwd(dir_with_FCS_files )
file.name <- "SPLN_L000108770_G02_015.labelled.fcs"
x <- read.FCS(file.name, alter.names=TRUE)
x

```

###Have a peek what is inside
Access raw events info and file metadata:
```{r, eval=FALSE}
summary(x)

exprs(x)
dim(exprs(x))

names(x)
featureNames(x)
colnames(x)
keyword(x)
description(x)
str(description(x))

#Compare reading with and without transformation (with summary())
x <- read.FCS(file.name)
y <- read.FCS(file.name, transformation="scale")
```

###Visualise

Library *ggcyto* builds on *ggplot2* to visualise flow cytometry. 


```{r, eval=FALSE}
require(ggcyto)

#One dimension
ggcyto(x, aes(x=APC.A)) +
  geom_density(fill="black")
autoplot(x, "APC.A")

#Two dimensions
ggcyto(x, aes(x=APC.A, y=PerCP.Cy5.5.A)) +
  geom_hex()
autoplot(x, "APC.A","PerCP.Cy5.5.A")

ggcyto(x, aes(x=APC.A, y=PerCP.Cy5.5.A)) + geom_hex(bins=128) + theme_bw()
autoplot(x, "APC.A","PerCP.Cy5.5.A", bins=128)

g <- ggcyto(x,aes(x = `FSC-H`, y =  `SSC-H`)) +
  geom_hex(bins=128)

#Might be usefulto limit plotting to the instrument range
g + ggcyto_par_set(limits = "instrument")

#Changing colour scale etc.
require(RColorBrewer)
g + scale_fill_gradientn(colours = brewer.pal(n=8,name="PiYG"),trans="sqrt")
g + scale_fill_gradient(trans = "sqrt", low = "gray", high = "black")
g + geom_density2d(colour = "black")


```


###Reading in a set of .FCS files

Usually however we want to read a set of FCS files - into a `flowSet`. `FlowSet` stores also metadata about the whole grup of files - like expemerimental covariates.
```{r, eval=FALSE}
fs <-read.flowSet(path=dir_with_FCS_files )
fs
colnames(fs)
pData(fs)
sampleNames(fs)
varLabels(fs)

name(fs)
pData(fs)$name <-paste("spleen", 1:6, sep="")
pData(fs)$genotype <-c(rep("KO",3), rep("WT",3))
fs[[1]]
fs[["spleen3"]]

dim(exprs(fs[["spleen3"]]))

autoplot(fs[[1]], 'FSC-A')
autoplot(fs, 'FSC-A') + ggcyto_par_set(limits = "instrument")
```


###Applying transformations and gating manually

*FlowCore* allows to explore effects of transformations and gating and split/subset `flowSet` and `flowFrame` accordingly. However, usually the `GatingHierarchy` and `GatingSet` are more suitable to use (scroll down).
```{r, eval=FALSE}
#(Advanced) play with transformations

#Define transformation function
mytransfFunction1 <- logTransform(logbase=10, r=1, d=1)

#Prepare a list with transformations and channels to which they have to be applied
myTrans <- transformList('FSC-A', mytransfFunction1) 

#Apply transformations from your list to the flowSet object
fs_f <-transform(fs, myTrans)

#inspect influenc of transformation
autoplot(fs[[1]], 'FSC-A')
autoplot(fs_f[[1]], 'FSC-A')
```

```{r, eval=FALSE}
#(Advanced) play with gates

#Define a rectangle gate
myRectGate <- rectangleGate(filterId="FluorescRegion", "FSC-H"=c(50000, 100000), "SSC-H"=c(0, 100000))

#flowCore allows also data-driven gates, like bivariate normal distribution gate
myNormGate <- norm2Filter("FSC-H", "SSC-H", method="covMcd", scale.factor=3, n=50000, filterId="defaultNorm2Filter")

fs_g1 <- filter(fs[[1]],myRectGate)
summary(fs_g1)

fs_g2 <- filter(fs[[1]],myNormGate)
summary(fs_g2)

#fs_g1 and fs_g2 are filterResult objects
```

```{r, eval=FALSE}
#(Advanced) Subset or split on gates
#
fs_3 <- Subset(fs, myRectGate)
fs_3
autoplot(fs_3, 'FSC-H')

fs_4 <- split(fs, myRectGate)
fs_4
autoplot(fs_4[[1]], 'FSC-H')
autoplot(fs_4[[2]], 'FSC-H')

#Gates might be overlaid/used together
filter(fs[[1]],myRectGate| myNormGate)
```



##GatingSet

`GatingSet` and `GatingHierarchy` - analogous to `flowSet` and `flowFrame` - store  data from a set/single FCS file along with transformations, gating and other metadata.
`GatingSet` migt be started by reading in a *FlowJo*/*Diva* Workspace, *OpenCyto* GatingML file or started fom a `flowSet` object. 
```{r, eval=FALSE}
require(flowWorkspace)

#to start a Gatingset from a flowSet
gs <- GatingSet(fs)
```


In the following example we'll use example files from the *flowWorkspace* vignette.

#####Initiate GatingSet
```{r, eval=FALSE }
#Read in the files into a GatingSet

  fcsfiles <- list.files(pattern = "CytoTrol", system.file("extdata", package = "flowWorkspaceData"), full = TRUE)
  fcsfiles
  fss <- read.flowSet(fcsfiles)
  
  #we want to change flowSet into gatingSet
  gs <- GatingSet(fss)
```


#####Compensate
```{r, eval=FALSE }
## ----Compensate----
# We'll cheat and apply a compensation matrix from another experiment. We'll read its compensation matrices
gs_manual <- load_gs(list.files(pattern = "gs_manual"
                                , system.file("extdata", package = "flowWorkspaceData")
                                , full = TRUE))
plot(gs_manual)
comp <- getCompensationMatrices(gs_manual[[1]])
comp

#appplying compensation
gs <- compensate(gs, comp)

```

#####Transform
```{r , eval=FALSE }

##We save un-transformed version of the data for future comparisons.
fs_comp <- getData(gs) 


#We define transformations - here logicle transformation for each of the parameters
##get all channels' names
chnls <- parameters(comp)
##estimate parameters for estimateLogicle directly on GatingHierarchy object to generate transformerList automatically
tf <- estimateLogicle(gs[[1]], chnls)
##apply transformation
gs <- transform(gs, tf)

## Plot
p1 <- autoplot(fs_comp[[1]], "B710-A") + ggtitle("raw")
p2 <- autoplot(flowData(gs)[[1]], "B710-A") + 
          ggtitle("trans") + 
          ggcyto_par_set(limits = "instrument")

require(gridExtra)
grid.arrange(as.ggplot(p1), as.ggplot(p2), ncol = 2)
```

#####Gate
```{r , eval=FALSE }

##Add a gate
rg1 <- rectangleGate("FSC-A"=c(50000, Inf), filterId="NonDebris")
add(gs, rg1, parent = "root") #!!!!!!it changes the object "gs" in situ!
getNodes(gs)
# gate the data
recompute(gs)

##  Plot
autoplot(gs, "NonDebris")

## Plot
ggcyto(gs, aes(x = `FSC-A`)) + geom_density() + geom_gate("NonDebris")  
```



#####Get stats
```{r , eval=FALSE }
getTotal(gs[[1]], "NonDebris")#counts
getProp(gs[[1]], "NonDebris")#proportion
getPopStats(gs)


```

To explore more, like combining different experiments, automated gating etc., read the *flowWorkspace* vignette.

##Working with other software derived files

###Reading in and visualising FlowJo workspaces

Requires a workspace (*.wsp* file) and FCS files.

```{r, eval=FALSE}
ws <- openWorkspace("MY PATH TO WHERE THIS FILE IS/MOEA T.wsp")
getSamples(ws)
getSampleGroups(ws)

gws <- parseWorkspace(ws,name=1)  
closeWorkspace(ws)

sampleNames(gws)


#Following part does not require starting from a FlowJo workspace, it will work with every GatingSet with appropriate data 
#Plot gating hierarchy:
plot(gws)

markernames(gws)
getData(gws[[1]])

getNodes(gws)
nodelist <- getNodes(gws, path = "auto")
node <- nodelist[3]

g <- getGate(gws, node)
g
getPopStats(gws)

#plot a chosen gate - this is a flowWorkspace function
plotGate(gws, "Tregs") #one gate for all samples

#add info about treatment
Genotype <- c("KO","WT","WT","KO","KO","WT","WT","WT","WT","KO","KO","KO")
pData(gws) <- cbind(pData(gws), Genotype)

subset(gws, Genotype == "KO")



gh <- gws[[1]]
head(getPopStats(gh))


pdf("sample1.pdf", width=15, height=20) #whole gating hierarchy
plotGate(gws[[1]])
dev.off()

#For more on working with flowJo wokspaces: 
vignette("plotGate")

```

### openCyto files
Requires an xml file and FCS files.
```{r, eval=FALSE}

#read in cytobank gatingML
fcsFiles <- list.files(pattern = "CytoTrol", system.file("extdata", package = "flowWorkspaceData"), full = T)
xmlfile <- system.file("extdata/cytotrol_tcell_cytobank.xml", package = "CytoML")
gs <- cytobank2GatingSet(xmlfile, fcsFiles)

```

###Diva Workspace
Requires an xml file and FCS files.
```{r, eval=FALSE}

#read in Diva workspace
ws <- openDiva(system.file('extdata/diva/PE_2.xml', package = "flowWorkspaceData"))
gs <- parseWorkspace(ws, name = 2, subset = 1)

```

###Importing a gating hierarchy (with transformations, compensation etc.)
Requires an xml file.
```{r , eval=FALSE}
#import gating hierarchy

library(CytoML)
xmlfile <- system.file("extdata/cytotrol_tcell_cytobank.xml", package = "CytoML")
g  <-read.gatingML.cytobank(xmlfile)
plot(g)

#now you can use gates and transformations from g on a GatingSet with the same channels:
trans <- getTransformations(g)
gs <- compensate(gs, g)
gs <- transform(gs, trans)
gating(g, gs) #gating acts in situ - this will change gs
```


